---
title: "Project proposal"
author: "hat team"
output: html_document
---

```{r load-packages, message = FALSE}
library(tidyverse)
library(broom)
library(readxl)
library(janitor)
library(forcats)
library(dplyr)
```

## 1. Introduction

Our data comes from the Promise Early Education Programs: Early Head Start and Head Start. They focus on children who come from low-income families and provide them with high quality early education. The program system uses GOLD assessment, evaluating the children through catergories of Physical Examination, Language, Cognitive, Social-Emotional, Literacy and Mathematics. For our project, we are focusing on the question of "What trends exist in social-emotional and behavioral development?" We will be using the data sets from both EHS and HS looking at the categories of Language, Cognitive and Literacy.

## 2. Data

Part of our longer term data analysis is figuring out how to wrangle this raw data and understand how to present it in a way that R can read. It is complicated and arranged in ways that are not suitable for R dataframes at this point.

```{r fall-data-cleanup}
fall_2018_19_data <- read_excel("../data/2018-19 promise data.xlsx", 
    sheet = "3-4", range = "A3:I81") |>
  janitor::clean_names() |>
  rename("category" = "x1")

fall_2018_19_data[1,1] <- "Social-Emotional"

fall_2018_19_data <- fall_2018_19_data |>
  filter(!category %in% c("Threes SE", "Threes")) 

fall_2018_19_data <- fall_2018_19_data |>
  fill(category, .direction = "down") |>
  drop_na(bottom)

fall_2018_19_data <- fall_2018_19_data |>
  filter(!bottom %in% c("Widely Held Expectations", "Bottom"))

fall_2018_19_data <- fall_2018_19_data |>
  mutate(season = "Fall") |>
  mutate(year = "18-19")

fall_2018_19_data$percent_meeting_exceeding <- as.numeric(fall_2018_19_data$percent_meeting_exceeding)

fall_2018_19_data$percent_below[1] <- 100
fall_2018_19_data$number_meeting_exceeding[1] <- 0

fall_2018_19_data <- fall_2018_19_data |>
  mutate(across(c(bottom, top, number_children, average, number_below, percent_below, number_meeting_exceeding, percent_meeting_exceeding), as.numeric))


fall_2018_19_data[is.na(fall_2018_19_data)] <- 0


fall_2018_19_mean <- fall_2018_19_data |>
  group_by(category) |>
  summarize_at(vars("percent_meeting_exceeding"), mean, na.rm = TRUE)
```

```{r winter-data-cleanup}
winter_2018_19_data <- read_excel("../data/2018-19 promise data.xlsx", 
    sheet = "3-4", range = "J3:O80")

expectations <- read_excel("../data/2018-19 promise data.xlsx", 
    sheet = "3-4", range = "A3:C80")


winter_2018_19_data <- cbind(expectations, winter_2018_19_data) |>
    janitor::clean_names()

winter_2018_19_data <- winter_2018_19_data |>
  rename(`category` = `x1`)

winter_2018_19_data[1, 1] <- "Social-Emotional"

winter_2018_19_data <- winter_2018_19_data |>
  filter(!category %in% c("Threes SE", "Threes")) 

winter_2018_19_data <- winter_2018_19_data |>
  fill(category, .direction = "down") |>
  drop_na(bottom)

winter_2018_19_data <- winter_2018_19_data |>
  filter(!bottom %in% c("Widely Held Expectations", "Bottom"))

winter_2018_19_data <- winter_2018_19_data |>
  mutate(season = "Winter") |>
   mutate(year = "18-19")

winter_2018_19_data$percent_meeting_exceeding <- as.numeric(winter_2018_19_data$percent_meeting_exceeding)

winter_2018_19_data <- winter_2018_19_data |>
  mutate(across(c(bottom, top, number_children, average, number_below, percent_below, number_meeting_exceeding, percent_meeting_exceeding), as.numeric))

winter_2018_19_data[is.na(winter_2018_19_data)] <- 0

winter_2018_19_mean <- winter_2018_19_data |>
  group_by(category) |>
  summarize_at(vars("percent_meeting_exceeding"), mean, na.rm = TRUE)
```

```{r spring-data-cleanup}
spring_2018_19_data <- read_excel("../data/2018-19 promise data.xlsx", 
    sheet = "3-4", range = "P3:U80")

spring_2018_19_data <- cbind(expectations, spring_2018_19_data) |>
    janitor::clean_names()


spring_2018_19_data <- spring_2018_19_data |>
  rename(`category` = `x1`)

spring_2018_19_data[1, 1] <- "Social-Emotional"


spring_2018_19_data <- spring_2018_19_data |>
  filter(!category %in% c("Threes SE", "Threes")) 

spring_2018_19_data <- spring_2018_19_data |>
  fill(category, .direction = "down") |>
  drop_na(bottom)

spring_2018_19_data <- spring_2018_19_data |>
  filter(!bottom %in% c("Widely Held Expectations", "Bottom"))

spring_2018_19_data <- spring_2018_19_data |>
  mutate(season = "Spring") |>
   mutate(year = "18-19")

spring_2018_19_data$percent_meeting_exceeding <- as.numeric(spring_2018_19_data$percent_meeting_exceeding)

spring_2018_19_data <- spring_2018_19_data |>
  mutate(across(c(bottom, top, number_children, average, number_below, percent_below, number_meeting_exceeding, percent_meeting_exceeding), as.numeric))

spring_2018_19_data[is.na(spring_2018_19_data)] <- 0

spring_2018_19_mean <- spring_2018_19_data |>
  group_by(category) |>
  summarize_at(vars("percent_meeting_exceeding"), mean, na.rm = TRUE)
```

```{r data-joining}
data_2018_19 <- rbind(fall_2018_19_data, winter_2018_19_data, spring_2018_19_data)

write.csv(data_2018_19, "../data/final_data.csv")
```

## 3. Data analysis plan

In our categories that we are analyzing, we will be looking at how the percentages of students below, meeting, and exceeding expectations changes throughout a year. We will also look to see if there are correlations between the growth/decline of students in multiple categories. We plan to analyze the difference in trends between the Early Head Start Program and the Head Start Program. We plan on doing this by using histograms, bar plots, and faceting them to compare different sets of data.

Our data still requires a lot of cleaning, so we do not have any preliminary visualizations. However, just by looking at the raw data, we notice some positive trends in students meeting/exceeding expectations throughout the school year.

```{r totals-by-season}
data_2018_19$season <- fct_relevel(data_2018_19$season, "Fall", "Winter", "Spring")

ggplot(data_2018_19, aes(x = season, y = percent_meeting_exceeding, group = category, color = category)) +
  stat_summary(fun = "mean", geom = "line", linewidth = 1.2) +  # Line for each category
  stat_summary(fun = "mean", geom = "point", linewidth = 3) +  # Points to highlight means
  labs(title = "Trend of Percent Meeting/Exceeding Expectations by Category", 
       subtitle = "Across 2018-19 Seasons", 
       x = "Season", 
       y = "Percent Meeting/Exceeding") +
  scale_color_brewer(palette = "Set2") +  # Aesthetic color palette
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1), 
        legend.title = element_blank())

```

## 4. Data Ethics Review

The data ethics review section will be introduced in a separate class and is not part of the original proposal deadline.
