---
title: "Project proposal"
author: "hat team"
output: html_document
---

```{r load-packages, message = FALSE}
library(tidyverse)
library(broom)
```

## 1. Introduction

Our data comes from the Promise Early Education Programs: Early Head Start and Head Start. They focus on children who come from low-income families and provide them with high quality early education. The program system uses GOLD assessment, evaluating the children through catergories of Physical Examination, Language, Cognitive, Social-Emotional, Literacy and Mathematics. 
For our project, we are focusing on the question of "What trends exist in social-emotional and behavioral development?" We will be using the data sets from both EHS and HS looking at the categories of Language, Cognitive and Literacy.  

## 2. Data

Text goes here.

```{r}
# Code goes here
```

## 3. Data analysis plan

Text goes here.


```{r}
# Code goes here
```


## 4. Data Ethics Review

The data ethics review section will be introduced in a separate class and is not part of the original proposal deadline.

