---
title: "Project proposal"
author: "hat team"
output: html_document
---

```{r load-packages, message = FALSE}
library(tidyverse)
library(broom)
library(readxl)
library(janitor)
library(forcats)
library(dplyr)
```

## 1. Introduction

Our data comes from the Promise Early Education Programs: Early Head Start and Head Start. They focus on children who come from low-income families and provide them with high quality early education. The program system uses GOLD assessment, evaluating the children through catergories of Physical Examination, Language, Cognitive, Social-Emotional, Literacy and Mathematics. For our project, we are focusing on the question of "What trends exist in social-emotional and behavioral development?" We will be using the data sets from both EHS and HS looking at the categories of Language, Cognitive and Literacy.

## 2. Data

Part of our longer term data analysis is figuring out how to wrangle this raw data and understand how to present it in a way that R can read. It is complicated and arranged in ways that are not suitable for R dataframes at this point.

```{r 18-19-fall-data-cleanup}
fall_2018_19_data <- read_excel("../data/2018-19 promise data.xlsx", 
    sheet = "3-4", range = "A3:I81") |>
  janitor::clean_names() |>
  rename("category" = "x1")

fall_2018_19_data[1,1] <- "Social-Emotional"

fall_2018_19_data <- fall_2018_19_data |>
  filter(!category %in% c("Threes SE", "Threes")) 

fall_2018_19_data <- fall_2018_19_data |>
  fill(category, .direction = "down") |>
  drop_na(bottom)

fall_2018_19_data <- fall_2018_19_data |>
  filter(!bottom %in% c("Widely Held Expectations", "Bottom"))

fall_2018_19_data <- fall_2018_19_data |>
  mutate(season = "Fall") |>
  mutate(year = "18-19") |>
  mutate(age_group = "3-4")

fall_2018_19_data$percent_meeting_exceeding <- as.numeric(fall_2018_19_data$percent_meeting_exceeding)

fall_2018_19_data$percent_below[1] <- 100
fall_2018_19_data$number_meeting_exceeding[1] <- 0

fall_2018_19_data <- fall_2018_19_data |>
  mutate(across(c(bottom, top, number_children, average, number_below, percent_below, number_meeting_exceeding, percent_meeting_exceeding), as.numeric))


fall_2018_19_data[is.na(fall_2018_19_data)] <- 0
```

```{r 18-19-winter-data-cleanup}
winter_2018_19_data <- read_excel("../data/2018-19 promise data.xlsx", 
    sheet = "3-4", range = "J3:O80")

expectations <- read_excel("../data/2018-19 promise data.xlsx", 
    sheet = "3-4", range = "A3:C80")


winter_2018_19_data <- cbind(expectations, winter_2018_19_data) |>
    janitor::clean_names()

winter_2018_19_data <- winter_2018_19_data |>
  rename(`category` = `x1`)

winter_2018_19_data[1, 1] <- "Social-Emotional"

winter_2018_19_data <- winter_2018_19_data |>
  filter(!category %in% c("Threes SE", "Threes")) 

winter_2018_19_data <- winter_2018_19_data |>
  fill(category, .direction = "down") |>
  drop_na(bottom)

winter_2018_19_data <- winter_2018_19_data |>
  filter(!bottom %in% c("Widely Held Expectations", "Bottom"))

winter_2018_19_data <- winter_2018_19_data |>
  mutate(season = "Winter") |>
   mutate(year = "18-19") |>
  mutate(age_group = "3-4")

winter_2018_19_data$percent_meeting_exceeding <- as.numeric(winter_2018_19_data$percent_meeting_exceeding)

winter_2018_19_data <- winter_2018_19_data |>
  mutate(across(c(bottom, top, number_children, average, number_below, percent_below, number_meeting_exceeding, percent_meeting_exceeding), as.numeric))

winter_2018_19_data[is.na(winter_2018_19_data)] <- 0
```

```{r 18-19-spring-data-cleanup}
spring_2018_19_data <- read_excel("../data/2018-19 promise data.xlsx", 
    sheet = "3-4", range = "P3:U80")

spring_2018_19_data <- cbind(expectations, spring_2018_19_data) |>
    janitor::clean_names()


spring_2018_19_data <- spring_2018_19_data |>
  rename(`category` = `x1`)

spring_2018_19_data[1, 1] <- "Social-Emotional"


spring_2018_19_data <- spring_2018_19_data |>
  filter(!category %in% c("Threes SE", "Threes")) 

spring_2018_19_data <- spring_2018_19_data |>
  fill(category, .direction = "down") |>
  drop_na(bottom)

spring_2018_19_data <- spring_2018_19_data |>
  filter(!bottom %in% c("Widely Held Expectations", "Bottom"))

spring_2018_19_data <- spring_2018_19_data |>
  mutate(season = "Spring") |>
   mutate(year = "18-19") |>
  mutate(age_group = "3-4")

spring_2018_19_data$percent_meeting_exceeding <- as.numeric(spring_2018_19_data$percent_meeting_exceeding)

spring_2018_19_data <- spring_2018_19_data |>
  mutate(across(c(bottom, top, number_children, average, number_below, percent_below, number_meeting_exceeding, percent_meeting_exceeding), as.numeric))

spring_2018_19_data[is.na(spring_2018_19_data)] <- 0
```

```{r 3-4-18-19-data-joining}
data_3_4_2018_19 <- rbind(fall_2018_19_data, winter_2018_19_data, spring_2018_19_data)
```

```{r 4-5-fall-data-cleanup}
fall_4_5_2018_19_data <- read_excel("../data/2018-19 promise data.xlsx", 
    sheet = "4-5", range = "A3:I109") |>
  janitor::clean_names() |>
  rename("category" = "x1")

fall_4_5_2018_19_data[1,1] <- "Social-Emotional"

fall_4_5_2018_19_data <- fall_4_5_2018_19_data |>
  filter(!category %in% c("Four to five", "Promise Early Education Center")) 

fall_4_5_2018_19_data <- fall_4_5_2018_19_data |>
  fill(category, .direction = "down") |>
  drop_na(bottom)

fall_4_5_2018_19_data <- fall_4_5_2018_19_data |>
  filter(!bottom %in% c("Widely Held Expectations", "Bottom"))

fall_4_5_2018_19_data <- fall_4_5_2018_19_data |>
  mutate(season = "Fall") |>
  mutate(year = "18-19") |>
  mutate(age_group = "4-5")

fall_4_5_2018_19_data$percent_meeting_exceeding <- as.numeric(fall_4_5_2018_19_data$percent_meeting_exceeding)

fall_4_5_2018_19_data$percent_below[1] <- 100
fall_4_5_2018_19_data$number_meeting_exceeding[1] <- 0

fall_4_5_2018_19_data <- fall_4_5_2018_19_data |>
  mutate(across(c(bottom, top, number_children, average, number_below, percent_below, number_meeting_exceeding, percent_meeting_exceeding), as.numeric))


fall_4_5_2018_19_data[is.na(fall_4_5_2018_19_data)] <- 0
```

```{r 4-5-winter-data-cleanup}
winter_4_5_2018_19_data <- read_excel("../data/2018-19 promise data.xlsx", 
    sheet = "4-5", range = "J3:O109")

expectations <- read_excel("../data/2018-19 promise data.xlsx", 
    sheet = "4-5", range = "A3:C109")


winter_4_5_2018_19_data <- cbind(expectations, winter_4_5_2018_19_data) |>
    janitor::clean_names()

winter_4_5_2018_19_data <- winter_4_5_2018_19_data |>
  rename(`category` = `x1`)

winter_4_5_2018_19_data[1,1] <- "Social-Emotional"

winter_4_5_2018_19_data <- winter_4_5_2018_19_data |>
  filter(!category %in% c("Four to five", "Promise Early Education Center")) 

winter_4_5_2018_19_data <- winter_4_5_2018_19_data |>
  fill(category, .direction = "down") |>
  drop_na(bottom)

winter_4_5_2018_19_data <- winter_4_5_2018_19_data |>
  filter(!bottom %in% c("Widely Held Expectations", "Bottom"))

winter_4_5_2018_19_data <- winter_4_5_2018_19_data |>
  mutate(season = "Winter") |>
  mutate(year = "18-19") |>
  mutate(age_group = "4-5")

winter_4_5_2018_19_data$percent_meeting_exceeding <- as.numeric(winter_4_5_2018_19_data$percent_meeting_exceeding)

winter_4_5_2018_19_data$percent_below[1] <- 100
winter_4_5_2018_19_data$number_meeting_exceeding[1] <- 0

winter_4_5_2018_19_data <- winter_4_5_2018_19_data |>
  mutate(across(c(bottom, top, number_children, average, number_below, percent_below, number_meeting_exceeding, percent_meeting_exceeding), as.numeric))


winter_4_5_2018_19_data[is.na(winter_4_5_2018_19_data)] <- 0
```

```{r 4-5-spring-data-cleanup}
spring_4_5_2018_19_data <- read_excel("../data/2018-19 promise data.xlsx", 
    sheet = "4-5", range = "P3:U109")

expectations <- read_excel("../data/2018-19 promise data.xlsx", 
    sheet = "4-5", range = "A3:C109")


spring_4_5_2018_19_data <- cbind(expectations, spring_4_5_2018_19_data) |>
    janitor::clean_names()

spring_4_5_2018_19_data <- spring_4_5_2018_19_data |>
  rename(`category` = `x1`)

spring_4_5_2018_19_data[1,1] <- "Social-Emotional"

spring_4_5_2018_19_data <- spring_4_5_2018_19_data |>
  filter(!category %in% c("Four to five", "Promise Early Education Center")) 

spring_4_5_2018_19_data <- spring_4_5_2018_19_data |>
  fill(category, .direction = "down") |>
  drop_na(bottom)

spring_4_5_2018_19_data <- spring_4_5_2018_19_data |>
  filter(!bottom %in% c("Widely Held Expectations", "Bottom"))

spring_4_5_2018_19_data <- spring_4_5_2018_19_data |>
  mutate(season = "Spring") |>
  mutate(year = "18-19") |>
  mutate(age_group = "4-5")

spring_4_5_2018_19_data$percent_meeting_exceeding <- as.numeric(spring_4_5_2018_19_data$percent_meeting_exceeding)

spring_4_5_2018_19_data$percent_below[1] <- 100
spring_4_5_2018_19_data$number_meeting_exceeding[1] <- 0

spring_4_5_2018_19_data <- spring_4_5_2018_19_data |>
  mutate(across(c(bottom, top, number_children, average, number_below, percent_below, number_meeting_exceeding, percent_meeting_exceeding), as.numeric))


spring_4_5_2018_19_data[is.na(spring_4_5_2018_19_data)] <- 0
```

```{r 4-5-18-19-data-joining}
data_4_5_2018_19 <- rbind(fall_4_5_2018_19_data, winter_4_5_2018_19_data, spring_4_5_2018_19_data)
```


```{r HS-2018-19-joining}
HS_2018_19_data <- rbind(data_3_4_2018_19, data_4_5_2018_19)

write.csv(HS_2018_19_data, "../data/HS_2018_19_data.csv")
```


## 3. Data analysis plan

In our categories that we are analyzing, we will be looking at how the percentages of students below, meeting, and exceeding expectations changes throughout a year. We will also look to see if there are correlations between the growth/decline of students in multiple categories. We plan to analyze the difference in trends between the Early Head Start Program and the Head Start Program. We plan on doing this by using histograms, bar plots, and faceting them to compare different sets of data.

Our data still requires a lot of cleaning, so we do not have any preliminary visualizations. However, just by looking at the raw data, we notice some positive trends in students meeting/exceeding expectations throughout the school year.

```{r means-by-season, fig.alt="Line plot showing the mean percent of 3–5 year olds meeting or exceeding expectations across Fall, Winter, and Spring 2018–19, broken down by developmental categories. Each category (such as Cognitive, Language, and Social-Emotional) is represented by a separate colored line using a colorblind-friendly viridis palette. Points mark the seasonal means for each category, showing overall upward trends across the year in most categories."}
HS_2018_19_data$season <- fct_relevel(HS_2018_19_data$season, "Fall", "Winter", "Spring")

ggplot(HS_2018_19_data, aes(x = season, y = percent_meeting_exceeding, group = category, color = category)) +
  stat_summary(fun = "mean", geom = "line", linewidth = 1.2) +  # Line for each category
  stat_summary(fun = "mean", geom = "point", linewidth = 3) +  # Points to highlight means
  labs(title = "3-5 Year Olds Percent Meeting/Exceeding Expectations", 
       subtitle = "All Categories Across 2018-19 Seasons", 
       x = "Season", 
       y = "Percent Meeting/Exceeding") +
  scale_color_viridis_d() + 
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1), 
        legend.title = element_blank())

ggsave("season_trend_2018_19.jpeg", path = "../memo")
```


```{r social-emotional-individuals, fig.alt="Scatterplot with a smooth LOESS trend line showing the percent of 3–5 year old students meeting or exceeding social-emotional expectations in Fall, Winter, and Spring of 2018–19. Individual group observations are plotted as viridis-colored dots. The trend increases slightly across the seasons."}
library(dplyr)
library(viridis)

# Filter for only the "Social-Emotional" category
social_emotional_data <- HS_2018_19_data %>% 
  filter(category == "Social-Emotional")

# Relevel the season factor
social_emotional_data$season <- fct_relevel(social_emotional_data$season, "Fall", "Winter", "Spring")

# Use viridis color
point_color <- viridis(1, option = "D")  # Option "D" is the default and colorblind safe

# Plot
ggplot(social_emotional_data, aes(x = season, y = percent_meeting_exceeding)) +
  geom_jitter(width = 0.2, size = 2, alpha = 0.7, color = point_color) +
  geom_smooth(aes(group = 1), method = "loess", se = TRUE, color = "black", fill = "grey70", linewidth = 1.2) +
  labs(title = "Group Observations of 3-5 Year Olds in Head Start Program", 
       subtitle = "Social-Emotional Category Across 2018-19 Seasons", 
       x = "Season", 
       y = "Percent Meeting/Exceeding") +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r HS-2018-19-SE-plot}
# Filter for only the "Social-Emotional" category
social_emotional_data <- HS_2018_19_data %>% 
  filter(category == "Social-Emotional")

# Relevel the season factor
social_emotional_data$season <- fct_relevel(social_emotional_data$season, "Fall", "Winter", "Spring")

# Plot with color by age group
ggplot(social_emotional_data, aes(x = season, y = percent_meeting_exceeding)) +
  geom_jitter(aes(color = age_group), width = 0.2, size = 2, alpha = 0.75) +
  geom_smooth(aes(group = 1), method = "loess", se = TRUE, color = "black", linewidth = 1.2) +
  labs(title = "Group Observations of Students in Head Start Program", 
       subtitle = "Social-Emotional Category Across 2018-19 Seasons by Age Group", 
       x = "Season", 
       y = "Percent Meeting/Exceeding",
       color = "Age Group") +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r line-plot, fig.alt="Line plot showing seasonal trends in social-emotional development scores for 3–4 and 4–5 year olds in the 2018–19 Head Start program. The x-axis shows seasons (Fall, Winter, Spring), and the y-axis shows the percent of students meeting or exceeding expectations. Two colored lines represent each age group using viridis colorblind-safe hues. Both groups show a general increase in scores across the year. The 3-4 age group shows higher social-emotional scores through all three seasons."}
# Filter for only the "Social-Emotional" category
social_emotional_data <- HS_2018_19_data %>%
  filter(category == "Social-Emotional")

# Relevel season factor
social_emotional_data$season <- fct_relevel(social_emotional_data$season, "Fall", "Winter", "Spring")

# --- Plot 1: Grouped Line Plot for Mean Trends ---
line_plot <- ggplot(social_emotional_data, aes(x = season, y = percent_meeting_exceeding, 
                                               group = age_group, color = age_group)) +
  stat_summary(fun = mean, geom = "line", linewidth = 1.2) +
  stat_summary(fun = mean, geom = "point", size = 3) +
  scale_color_viridis_d(option = "D") +
  labs(title = "Trend of Head Start Program Social-Emotional Scores",
       subtitle = "Mean Percent Meeting/Exceeding Expectations (2018-19) by Season and Age Group",
       x = "Season", y = "Percent Meeting/Exceeding",
       color = "Age Group") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave("2018_19_3_4_vs_4_5_line_plot.jpeg", path = "../memo")

line_plot

```
```{r box-plot, fig.alt="Faceted boxplot showing the distribution of social-emotional scores by age group across Fall, Winter, and Spring of 2018–19. The x-axis shows age groups (3–4 and 4–5), and the y-axis shows percent meeting or exceeding expectations. Each panel represents a season. Boxplots use viridis colors and show central tendency and variability within each group-season pair. Once again, the 3-4 age group shows higher social-emotional scores overall."}

# Filter for only the "Social-Emotional" category
social_emotional_data <- HS_2018_19_data %>%
  filter(category == "Social-Emotional")

# Relevel season factor
social_emotional_data$season <- fct_relevel(social_emotional_data$season, "Fall", "Winter", "Spring")

# --- Plot 2: Faceted Boxplot with Jittered Points ---
box_jitter_plot <- ggplot(social_emotional_data, aes(x = age_group, y = percent_meeting_exceeding, fill = age_group)) +
  geom_boxplot(width = 0.5, outlier.shape = NA, alpha = 0.8) +
  facet_wrap(~ season) +
  scale_fill_viridis_d(option = "D") +
  labs(title = "Distribution of Head Start Program Social-Emotional Scores",
       subtitle = "Percent Meeting/Exceeding Expectations (2018-19) by Season and Age Group",
       x = "Age Group", y = "Percent Meeting/Exceeding",
       fill = "Age Group") +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1))

ggsave("2018_19_boxplot_seasons.jpeg", path = "../memo")

box_jitter_plot
```

## 4. Data Ethics Review

The data ethics review section will be introduced in a separate class and is not part of the original proposal deadline.
